---
title: 2020 First Presidential Debate Analysis
author: Srikar Durgam
date: '2020-10-06'
slug: first-presidetial-debate-analysis
categories: []
tags:
  - debate
  - president
  - tutorial
---

My first blog post! If you don't know me already, hello! My name is Srikar. I am a data analyst for an ISP in Atlanta, GA and currently in the middle of getting my business analytics and data science masters degree from Wake Forest. 

The purpose of this blog post isn't to nudge you in one way or another, but only to provide an analytical view of what the candidates are saying. Please treat my blog and skills as a work in progress. I have been working with R for over a year now and I am still learning new things about it. I truly have a passion for data and love the ability to have an opportunity to tell a story using data. Feel free to use my git repository (links are coming soon) to conduct your analysis!

First we are going to load our libraries. We have the usual suspects here for text analysis.
```{r Library, echo=TRUE, message=FALSE, warning=FALSE}
library(tidyverse)
library(tidytext)
library(ggplot2)
library(readr)
library(wordcloud2)
library(janeaustenr)
library(dplyr)
library(stringr)
```

I was able to find a simple data set with three columns(speaker, time and test) in Kaggle, the same data set can be found  [here](https://www.kaggle.com/headsortails/us-election-2020-presidential-debates). 
We then import the dataset. I also converted speaker column to factor data type to make it easier to model at a later time. I then check for the structure of the data using the str command to see if I sucessfully converted the speaker column to the factor data type. 

```{r message=FALSE, warning=FALSE}
presidential_debate <- read_csv("us_election_2020_1st_presidential_debate.csv", 
    col_types = cols(minute = col_skip(), 
        speaker = col_factor(levels = c("Chris Wallace", 
            "President Donald J. Trump", 
            "Vice President Joe Biden"))))
```

Lastly, because we are mainly interested in understanding what the candidates are saying I decided to filter out all the columns with the host as the speaker (Chris Wallace). Then I print out the first 5 columns in the dataset.

```{r message=FALSE, warning=FALSE}
presidential_debate <- presidential_debate %>%
                       filter(!speaker %in% c('Chris Wallace'))

head(presidential_debate)
```

Initially my goal was to create a word cloud and then analyze those words based on the sentiment. So at the end of this analysis we should have two word clouds(one for each candidate) and 4 charts that illustrate positive and negative words, given the context of the word. Because there were a lot of words I decided to only look at the top 100 words used by each candidate. 

First up, its President Donald Trump!

```{r message=FALSE, warning=FALSE}
DJT <- presidential_debate %>% 
       unnest_tokens(word, "text", token = "ngrams", n =1) %>%
       anti_join(stop_words) %>%
       filter(speaker == 'President Donald J. Trump') %>%
       group_by(word) %>% 
       summarise(n=n()) %>%
       arrange(desc(n)) %>%
       top_n(100,n)

DJT %>% wordcloud2()
```


His most frequent word during the first debate was "people". The "crosstalk" we see here is them interrupting each other, I am sure this will show up in the Joe Biden word cloud as well. We can see that President Trump spoke a lot about VP Joe Biden, the left, the ballot/ballots, the country, the military, Obamacare and the forest. After this initial view I think there is some words we need to filter like the words, "don't, it's, you've, you're ". etc

```{r message=FALSE, warning=FALSE}
DJT <- presidential_debate %>% 
       unnest_tokens(word, "text", token = "ngrams", n =1) %>%
       anti_join(stop_words) %>%
       filter(speaker == 'President Donald J. Trump') %>%
       filter(!word %in% c("crosstalk", "it’s","don’t", "that’s", "i’m", "didn’t", "they’re", "00", "you’re",
                           "we’re", "we’ve", "i’ll","he’s", "i’ve", "you’ve")) %>%
       group_by(word) %>% 
       
       summarise(n=n()) %>%
       arrange(desc(n)) %>%
       top_n(100,n)

DJT %>% wordcloud2()
```


I think the second wordcloud is a lot more cleaner and also conveys some of the important topics that President Trump discussed during this debate. 

Now lets do the same for VP Joe Biden. For the sake saving time and being fair I will filter the same words as I did for President Trump. 

```{r message=FALSE, warning=FALSE}
Joe <- presidential_debate %>% 
       unnest_tokens(word, "text", token = "ngrams", n =1) %>%
       anti_join(stop_words) %>%
       filter(speaker == 'Vice President Joe Biden') %>%
       filter(!word %in% c("crosstalk", "it’s","don’t", "that’s", "i’m", "didn’t", "they’re", "00", "you’re",
                           "we’re", "we’ve", "i’ll","he’s", "i’ve", "you’ve")) %>%
       group_by(word) %>% 
       summarise(n=n()) %>%
       arrange(desc(n)) %>%
       top_n(100,n)

Joe %>% wordcloud2()
```

Looks like the word "people'' seems to be a popular one amongst the both candidates. VP Biden also talked about healthcare, police, vaccines, president(referring to President Trump I'd assume). One word that was mentioned in VP Biden's word cloud and not in President Trump word cloud is Covid. President Trump does mention China.

Now onto sentiment, lets look at words amongst both the candidates used in postive and negative connotation.

```{r message=FALSE, warning=FALSE}
library(reshape2)

DJTSentiment <- DJT %>% inner_join(get_sentiments("bing"), by = "word") %>%
                        group_by(sentiment) %>%
                        top_n(10,n) %>%
                        mutate(score = if_else(sentiment == "positive", n, -n))

DJTSentiment %>%      ggplot(aes(reorder(word,n),n , fill = sentiment)) + 
                      geom_col() + 
                      coord_flip() + 
                      facet_wrap(~ sentiment, scales = "free_y") + 
                      labs(title = "Top 10 Terms by Sentiment: President Donald Trump",
                         x = "term", y = "score")
```

President Trump used his own name in a positive way which I think is very intersting. I think his use of vice in the negative terms probably refers to Vice President Joe Biden. 

```{r message=FALSE, warning=FALSE}
JoeSentiment <- Joe %>% inner_join(get_sentiments("bing"), by = "word") %>%
                        group_by(sentiment) %>%
                        top_n(10,n) %>%
                        mutate(score = if_else(sentiment == "positive", n, -n))

JoeSentiment %>%      ggplot(aes(reorder(word,n),n , fill = sentiment)) + 
                      geom_col() + 
                      coord_flip() + 
                      facet_wrap(~ sentiment, scales = "free_y") + 
                      labs(title = "Top 10 Terms by Sentiment: Vice President Joe Biden",
                         x = "term", y = "score")
```

I am not really sure why the word vice is showing up on VP Biden's speech as well but I think he is probably referring to VP Pence, but that is just speculation. Other than that we have the usual suspects here as well.

In conclusion, I used text analytics techniques to analyze the first presidential debate. I will probably do the same for the remainder of the debates. Please be sure to vote in this year's election to make your voices heard. Another metric we should look at in the next debate is who is spending the most time talking. We can then plot that against the alloted time for each candidate to see who speaking the most and are they going over their allotted time, this will also give us an idea of how well the debate moderator performed. 

Thank you for taking the time and reading my blog. 


